---
title: "ML Final Project"
author: 'Sabrina Wu, Mandy Pu'
date: "2026-05-08"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# When this report is knitted from the analysis/ folder, use the repository root
# as the working directory so data paths remain stable.
knitr::opts_knit$set(root.dir = normalizePath(file.path(getwd(), ".."), mustWork = TRUE))
```

# Load Packages 

```{r, warning=FALSE, results = 'hide'}
library(forcats)   # releveling factors
library(xgboost)   # xgb.DMatrix, xgb.cv, xgb.train, xgb.importance
library(caret)     # createDataPartition, confusionMatrix
library(pROC)      # roc, auc
library(corrplot)  # correlation heatmap
library(tidyverse) # data manipulation
library(kernlab)   # ksvm
library(gridExtra) # grid.arrange
library(ggplot2)   # visualization
library(gtsummary) # tables
```

# Data Preparation

Data Source: UCI ML Repository, https://archive.ics.uci.edu/dataset/45/heart+disease

This is a database on heart diseases. 

The database contains 4 processed files that contain the following features:
age, sex, cp, trestbps, chol, fbs, restecg, thalach, exang, oldpeak, slope, ca, thal, num

Target: num — 0 = no disease, 1–4 = disease present (binarized later)

## 1. Load datasets

```{r}
col_names <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg",
               "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")

# helper function to read dataset
read_hd <- function(path) {
  read.csv(path, header = FALSE, col.names = col_names,
           na.strings = "?", stringsAsFactors = FALSE)
}

# load all 4 datasets
cleveland   <- read_hd("data/raw/heart_disease/processed.cleveland.data")
hungarian   <- read_hd("data/raw/heart_disease/processed.hungarian.data")
switzerland <- read_hd("data/raw/heart_disease/processed.switzerland.data")
va          <- read_hd("data/raw/heart_disease/processed.va.data")

datasets <- list(
  cleveland   = cleveland,
  hungarian   = hungarian,
  switzerland = switzerland,
  va          = va
)
```

## 2. Missingness Check

```{r}
# helper func to calculate missingness percentage
miss_pct <- function(df) round(100 * colSums(is.na(df)) / nrow(df), 1)

miss_table <- do.call(rbind, lapply(names(datasets), function(nm) {
  pct <- miss_pct(datasets[[nm]])
  row <- as.data.frame(t(pct))
  row$site          <- nm
  row$n_rows        <- nrow(datasets[[nm]])
  row$complete_rows <- sum(complete.cases(datasets[[nm]]))
  row$pct_complete  <- round(100 * row$complete_rows / row$n_rows, 1)
  row
}))
rownames(miss_table) <- NULL

# Print the missingness table — site overview first
overview <- miss_table[, c("site", "n_rows", "complete_rows", "pct_complete")]
print(overview, row.names = FALSE)

# column wise missingness percentage
print(miss_table[, c("site", col_names)], row.names = FALSE)

# Highlight columns with high structural missingness across non-Cleveland sites
non_clev <- miss_table[miss_table$site != "cleveland", c("site", col_names)]
high_miss <- names(which(colMeans(non_clev[, col_names]) > 30))
high_miss
```

In the missingness analysis, `slope`, `ca`, and `thal` features had missingness >30% in all datasets except for Cleveland. The dataset completeness for hungarian, switzerland, and va datasets are low, almost close to 0% due to the large percentage of missing values in these three features. This is consistent with the description on the data source page. Therefore, the following analysis will be done on the cleveland dataset only. 

Below is the missingness check on cleveland dataset only: 

```{r}
clev_miss <- data.frame(
  column   = col_names,
  n_miss   = colSums(is.na(cleveland)),
  pct_miss = miss_pct(cleveland),
  row.names = NULL
)
print(clev_miss[clev_miss$n_miss > 0, ], row.names = FALSE)
cat(sprintf("\nComplete rows: %d / %d  (%.1f%% complete)\n",
            sum(complete.cases(cleveland)), nrow(cleveland),
            100 * mean(complete.cases(cleveland))))
```

Feature `ca` has 4 NA values and `thal` has 2 NA values. Given the low missing value percentage, we will drop these 6 rows from analysis. 

## 3. Validity Check 

To check whether all values are valid in the dataset. 

```{r}
impossible <- data.frame(
  check         = c("trestbps == 0", "chol == 0", "oldpeak < 0",
                    "age < 18 or > 100", "thalach == 0"),
  n_flagged     = c(
    sum(!is.na(cleveland$trestbps) & cleveland$trestbps == 0),
    sum(!is.na(cleveland$chol)     & cleveland$chol     == 0),
    sum(!is.na(cleveland$oldpeak)  & cleveland$oldpeak  <  0),
    sum(!is.na(cleveland$age)      & (cleveland$age < 18 | cleveland$age > 100)),
    sum(!is.na(cleveland$thalach)  & cleveland$thalach  == 0)
  )
)
print(impossible, row.names = FALSE)
```

No impossible or invalid values found in the dataset.

# Data Cleaning 
```{r}
# Drop incomplete rows (ca and thal missings)
cleveland_clean <- cleveland[complete.cases(cleveland), ]

# Binarize target: 0 = no disease, otherwise 1
cleveland_clean$target <- factor(
  ifelse(cleveland_clean$num == 0, 0, 1),
  levels = c(0, 1), labels = c("no_disease", "disease")
)
cleveland_clean$num <- NULL 

# Recode categorical predictors with more than 2 categories as factors
cleveland_clean$cp      <- factor(cleveland_clean$cp,
                                  levels = 1:4,
                                  labels = c("typical_angina", "atypical_angina",
                                             "non_anginal", "asymptomatic")) 

# reordered the levels so that asymptomatic is the reference group in models
cleveland_clean$cp <- fct_relevel(cleveland_clean$cp, 
                                  "asymptomatic", 
                                  "non_anginal", 
                                  "atypical_angina", 
                                  "typical_angina") 
 
cleveland_clean$restecg <- factor(cleveland_clean$restecg,
                                  levels = c(0, 1, 2),
                                  labels = c("normal", "ST-T wave abnormality", "hypertrophy"))

cleveland_clean$slope   <- factor(cleveland_clean$slope,
                                  levels = 1:3,
                                  labels = c("upsloping", "flat", "downsloping"))

cleveland_clean$thal    <- factor(cleveland_clean$thal,
                                  levels = c(3, 6, 7),
                                  labels = c("normal", "fixed_defect",
                                             "reversible_defect"))

head(cleveland_clean)

str(cleveland_clean)
```

The final cleaned dataset has 297 rows and 14 variables, which means 13 features and one target. 

# EDA

Summary statistics
```{r}
df <- cleveland_clean
summary(df)
```

Disease prevalence 
```{r}
prop.table(table(df$target))
```

Visualizations: 
```{r}
p1 <- ggplot(df, aes(x = target, fill = target)) +
geom_bar(color = "black", alpha = 0.8) +
theme_minimal() +
labs(title = "Figure: 1: Distribution of Heart Disease (Target)", x = "Diagnosis", y = "Count") +
scale_fill_manual(values = c("#E69F00", "#56B4E9"))
print(p1)
```

```{r}
p2 <- ggplot(df, aes(x = target, y = age, fill = target)) +
geom_boxplot(alpha = 0.7) +
theme_minimal() +
labs(title = "Figure 2: Age Distribution by Heart Disease Status", x = "Diagnosis", y = "Age") +
scale_fill_manual(values = c("#E69F00", "#56B4E9"))
print(p2)
```

```{r}
p3 <- ggplot(df, aes(x = thalach, fill = target)) +
geom_density(alpha = 0.5) +
theme_minimal() +
labs(title = "Figure 3: Max Heart Rate Density by Diagnosis", x = "Max Heart Rate (thalach)", y = "Density")
print(p3)
```

Univariable table
```{r}
univariable_table <- df %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    label = list(
      age ~ "Age (years)",
      sex ~ "Sex",
      trestbps ~ "Resting blood pressure (mm Hg)",
      chol ~ "Serum cholesterol (mg/dl)",
      thalach ~ "Maximum heart rate achieved",
      oldpeak ~ "ST depression (Induced by Exercise)",
      fbs ~ "Fasting blood sugar > 120 mg/dl",
      thal ~ "Thallium stress test",
      cp ~ "Chest pain type",
      restecg ~ "Resting electrocardiographic results",
      exang ~ "Exercise induced angina",
      slope ~ "The slope of the peak exercise ST segment",
      ca ~ "Number of major vessels (0-3) colored by flourosopy",
      target ~ "Heart disease diagnosis"
      )
  ) %>%
  bold_labels()
univariable_table

#Bivariable Analysis
df_bivariate <- df %>%
  mutate(
    target = factor(target, levels = c("no_disease", "disease"), 
                    labels = c("No Heart Disease", "Heart Disease"))
  )
```

Bivariable table

```{r}
bivariable_table <- df_bivariate %>%
  tbl_summary(
    by = target, 
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany",
    label = list(
      age ~ "Age (years)",
      sex ~ "Sex",
      trestbps ~ "Resting blood pressure (mm Hg)",
      chol ~ "Serum cholesterol (mg/dl)",
      thalach ~ "Maximum heart rate achieved",
      oldpeak ~ "ST depression (Induced by Exercise)",
      fbs ~ "Fasting blood sugar > 120 mg/dl",
      thal ~ "Thallium stress test",
      cp ~ "Chest pain type",
      restecg ~ "Resting electrocardiographic results",
      exang ~ "Exercise induced angina",
      slope ~ "The slope of the peak exercise ST segment",
      ca ~ "Number of major vessels (0-3) colored by flourosopy"
      )
  ) %>%
  add_overall() %>%
  add_p() %>%       
  bold_labels()

bivariable_table
```

Correlation Matrix

```{r}
numeric_vars <- df %>%
  select(where(is.numeric)) %>%
  rename(
    `Age (years)` = age,
    `Resting BP (mm Hg)` = trestbps,
    `Cholesterol (mg/dl)` = chol,
    `Max Heart Rate` = thalach,
    `ST Depression` = oldpeak)

cor_matrix <- cor(numeric_vars)

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 0.8,
         addCoef.col = "black",
         title = "Correlation Matrix of Numeric Features",
         mar = c(0,0,1,0))
```


# Model Building Preparation

Split dataset into 80% training and 20% testing while preserving the same proportion of binary disease classification in the entire dataset.

```{r}
set.seed(42)
train_idx <- createDataPartition(df$target, p = 0.80, list = FALSE)

df_train <- df[ train_idx, ]
df_test  <- df[-train_idx, ]

cat(sprintf("Training set : %d rows\n", nrow(df_train)))
cat(sprintf("Test set     : %d rows\n", nrow(df_test)))
```

Training set : 238 rows
Test set     : 59 rows

# Data Preprocessing

## 1. Scale continuous variables 
```{r}
cont_vars <- c("age", "trestbps", "chol", "thalach", "oldpeak")

train_means <- colMeans(df_train[, cont_vars])
train_sds   <- apply(df_train[, cont_vars], 2, sd)

scale_transform <- function(df_in) {
  df_out <- df_in
  for (v in cont_vars) {
    df_out[[v]] <- (df_in[[v]] - train_means[v]) / train_sds[v]
  }
  df_out
}

df_train_sc <- scale_transform(df_train)
df_test_sc  <- scale_transform(df_test)

head(df_train_sc)
head(df_test_sc)
```

## 2. Create dummy variables

Dummy-encode remaining factors - needed for XGBoost and SVM.
Use ~ . (with intercept) then drop the intercept column [, -1] so that every factor consistently loses exactly one reference level:

cp       → asymptomatic dropped 

restecg  → normal dropped        

slope    → upsloping dropped    

thal     → normal dropped       

```{r}
X_train <- model.matrix(target ~ ., data = df_train_sc)[, -1]
X_test  <- model.matrix(target ~ ., data = df_test_sc)[, -1]
```

## Target 

```{r}
y_train <- as.integer(df_train$target == "disease")
y_test  <- as.integer(df_test$target  == "disease")
```


# Model Fitting 

## 1. Benchmark - logistic regression

We explored both the full logistic regression and stepwise selection. 

Model 1: Full Logistic regression 

Since logistic regression handles categorical variables internally, we use the scaled training set 
```{r}
# Fit logistic regression on scaled training data (glm handles factors natively)
lr_model <- glm(target ~ ., data = df_train_sc, family = "binomial")
summary(lr_model)
exp(cbind(OR = coef(lr_model), confint(lr_model)))
```

Use Youden's J to find the optimal threshold for classification

```{r}
# Find optimal threshold via Youden's J on training ROC
lr_train_prob <- predict(lr_model, type = "response")
roc_lr        <- roc(df_train_sc$target, lr_train_prob, levels = c("no_disease", "disease"))
lr_coords     <- coords(roc_lr, "best", best.method = "youden",
                        ret = c("threshold", "sensitivity", "specificity"))
lr_thresh     <- lr_coords$threshold
cat(sprintf("Optimal threshold (Youden's J): %.4f\n", lr_thresh))
```

Optimal threshold (Youden's J): 0.4106

```{r}
# Train error
lr_train_label <- ifelse(lr_train_prob >= lr_thresh, "disease", "no_disease")
lr_train_err   <- mean(lr_train_label != as.character(df_train_sc$target))
cat(sprintf("Train error: %.4f\n", lr_train_err))

# Test error
lr_test_prob  <- predict(lr_model, newdata = df_test_sc, type = "response")
lr_test_label <- ifelse(lr_test_prob >= lr_thresh, "disease", "no_disease")
lr_test_err   <- mean(lr_test_label != as.character(df_test_sc$target))
cat(sprintf("Test error : %.4f\n", lr_test_err))
```

Train error: 0.1176; Test error : 0.1525

Evaluation:

```{r}
lr_preds <- factor(lr_test_label, levels = c("no_disease", "disease"))
lr_cm    <- confusionMatrix(lr_preds, df_test_sc$target, positive = "disease")
print(lr_cm)

lr_roc <- roc(df_test_sc$target, lr_test_prob, levels = c("no_disease", "disease"), quiet = TRUE)
cat(sprintf("Test AUC: %.4f\n", auc(lr_roc)))
lr_f1 <- lr_cm$byClass["F1"]
cat(sprintf("F1 score: %.4f\n", lr_f1))
```

Confusion matrix Visualization: 
```{r}
# confusion matrix visualization
lr_cm_df <- as.data.frame(lr_cm$table)
print(
  ggplot(lr_cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 7, fontface = "bold") +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "Count") +
    labs(title    = "Logistic Regression — Confusion Matrix",
         x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 13)
)
```

ROC Curve Visualization
```{r}
print(
  ggplot(data.frame(tpr = lr_roc$sensitivities,
                    fpr = 1 - lr_roc$specificities),
         aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    labs(title = "Logistic Regression — ROC Curve", x = "False Positive Rate", y = "True Positive Rate") +
    theme_minimal()
)
```


Model 2: Stepwise selection
```{r}
step_log_model <- step(lr_model, direction = "both", trace = 0)
summary(step_log_model)
```

```{r}
# Find optimal threshold via Youden's J on training ROC
step_train_prob <- predict(step_log_model, type = "response")
roc_step        <- roc(df_train_sc$target, step_train_prob, levels = c("no_disease", "disease"))
step_coords     <- coords(roc_step, "best", best.method = "youden",
                        ret = c("threshold", "sensitivity", "specificity"))
step_thresh     <- step_coords$threshold
cat(sprintf("Optimal threshold (Youden's J): %.4f\n", step_thresh))
```

Optimal threshold (Youden's J): 0.4754

```{r}
# Train error
step_train_label <- ifelse(step_train_prob >= step_thresh, "disease", "no_disease")
step_train_err   <- mean(step_train_label != as.character(df_train_sc$target))
cat(sprintf("Train error: %.4f\n", step_train_err))

# Test error
step_test_prob  <- predict(step_log_model, newdata = df_test_sc, type = "response")
step_test_label <- ifelse(step_test_prob >= step_thresh, "disease", "no_disease")
step_test_err   <- mean(step_test_label != as.character(df_test_sc$target))
cat(sprintf("Test error : %.4f\n", step_test_err))
```

Train error: 0.1218; Test error : 0.2034

Evaluation:

```{r}
step_preds <- factor(step_test_label, levels = c("no_disease", "disease"))
step_cm    <- confusionMatrix(step_preds, df_test_sc$target, positive = "disease")
print(step_cm)

step_roc <- roc(df_test_sc$target, step_test_prob, levels = c("no_disease", "disease"), quiet = TRUE)
cat(sprintf("Test AUC: %.4f\n", auc(step_roc)))
step_f1 <- step_cm$byClass["F1"]
cat(sprintf("F1 score: %.4f\n", step_f1))
```

Confusion matrix Visualization: 
```{r}
# confusion matrix visualization
step_cm_df <- as.data.frame(step_cm$table)
print(
  ggplot(step_cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 7, fontface = "bold") +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "Count") +
    labs(title    = "Logistic Regression (stepwise) — Confusion Matrix",
         x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 13)
)
```

Since the stepwise model has lower accuracy and sensitivity, we did not include it in the final analysis report. We focused on the full logistic regression model, SVM, and XGBoost. 


## 2. SVM

Use 5-fold CV and grid search method combined to find the best combination of Cost and gamma hyperparameters that gives the highest CV AUC. 

```{r}
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# all possible combinations of hyperparameters
svm_grid <- expand.grid(
  C     = c(0.1, 1, 10, 100),
  sigma = c(0.001, 0.01, 0.1, 1)
)

set.seed(42)
svm_model <- train(
  x         = X_train,
  y         = df_train$target,
  method    = "svmRadial",
  trControl = cv_ctrl,
  tuneGrid  = svm_grid,
  metric    = "ROC"
)
```

A visualization of the grid search method. 

```{r}
# Grid search heatmap — CV AUC across all C × sigma combinations
svm_res        <- svm_model$results
svm_res$C      <- factor(svm_res$C)
svm_res$sigma  <- factor(svm_res$sigma)

print(
  ggplot(svm_res, aes(x = C, y = sigma, fill = ROC)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.3f", ROC)), size = 3.5) +
    scale_fill_gradient(low = "lightyellow", high = "steelblue",
                        name = "CV AUC") +
    geom_tile(data = subset(svm_res,
                             C     == as.character(svm_model$bestTune$C) &
                             sigma == as.character(svm_model$bestTune$sigma)),
              color = "red", fill = NA, linewidth = 1.2) +
    labs(title    = "SVM Grid Search — 5-Fold CV AUC",
         subtitle = sprintf("Best: C = %s, gamma = %s",
                            svm_model$bestTune$C, svm_model$bestTune$sigma),
         x = "C (cost)", y = "Gamma") +
    theme_minimal()
)

```

Based on the grid search result, the model is fitted using C = 1 and gamma = 0.001. 

Next, we find the optimal classification threshold using Youden's J statistic. 

```{r}
# Get predicted probabilities for training data
pred_train_prob <- predict(svm_model, newdata = X_train, type = "prob")[, "disease"]

# Find optimal threshold via Youden's J on training ROC
roc_train   <- roc(df_train$target, pred_train_prob, levels = c("no_disease", "disease"))
best_coords <- coords(roc_train, "best", best.method = "youden",
                      ret = c("threshold", "sensitivity", "specificity"))
opt_thresh  <- best_coords$threshold
cat(sprintf("Optimal threshold (Youden's J): %.4f\n", opt_thresh))
```

The optimal J statistic is 0.5639. Using this threshold to calculate train and test error.

```{r}
# Train error using optimal threshold
pred_train_label <- ifelse(pred_train_prob >= opt_thresh, "disease", "no_disease")
train_err <- mean(pred_train_label != as.character(df_train$target))
cat(sprintf("Train error: %.4f\n", train_err))

# Test error using the same threshold
svm_probs <- predict(svm_model, newdata = X_test, type = "prob")[, "disease"]
svm_preds <- ifelse(svm_probs >= opt_thresh, "disease", "no_disease")
test_err  <- mean(svm_preds != as.character(df_test$target))
cat(sprintf("Test error: %.4f\n", test_err))
```

Train error = 0.1513 and test error = 0.1186.

Evaluation:
```{r}
# convert to factor for confusionMatrix downstream
svm_preds <- factor(svm_preds, levels = c("no_disease", "disease"))
# confusion matrix
svm_cm <- confusionMatrix(svm_preds, df_test$target, positive = "disease")
print(svm_cm)

svm_roc <- roc(y_test, svm_probs, quiet = TRUE)
cat(sprintf("Test AUC: %.4f\n", auc(svm_roc)))
svm_f1 <- svm_cm$byClass["F1"]
cat(sprintf("F1 score: %.4f\n", svm_f1))
```

Confusion matrix Visualization: 
```{r}
# confusion matrix visualization
svm_cm_df <- as.data.frame(svm_cm$table)
print(
  ggplot(svm_cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 7, fontface = "bold") +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "Count") +
    labs(title    = "SVM — Confusion Matrix",
         x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 13)
)
```

ROC Curve Visualization
```{r}
print(
  ggplot(data.frame(tpr = svm_roc$sensitivities,
                    fpr = 1 - svm_roc$specificities),
         aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    labs(title = "SVM — ROC Curve", x = "False Positive Rate", y = "True Positive Rate") +
    theme_minimal()
)
```

## 3. XGBoost

Create XGBoost Matrix objects
```{r}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)
```

Using 5-fold CV to finding the best tune hyperparameters that give the highest CV AUC.

Phase 1 -- Tree structure : max_depth, min_child_weight

Phase 2 -- Sampling       : subsample, colsample_bytree

Phase 3 -- Regularization : gamma

Phase 4 -- Learning rate  : eta 

```{r}
# A helper runs 5-fold CV for a given params list and returns the best AUC and the optimal nrounds (via early stopping).

run_cv <- function(params, nrounds = 300, early_stop = 20) {
  set.seed(42)
  cv <- xgb.cv(
    params                = params,
    data                  = dtrain,
    nrounds               = nrounds,
    nfold                 = 5,
    early_stopping_rounds = early_stop,
    verbose               = 0,
    maximize              = TRUE
  )
  list(
    best_nrounds = cv$best_iteration,
    best_auc     = cv$evaluation_log$test_auc_mean[cv$best_iteration]
  )
}

base_params <- list(
  booster          = "gbtree",
  objective        = "binary:logistic",
  eval_metric      = "auc",
  eta              = 0.1,
  max_depth        = 3,
  min_child_weight = 1,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  gamma            = 0
)
```

Phase 1: Start with max_depth and min_child_weight

```{r}
phase1_grid    <- expand.grid(max_depth        = c(3, 4, 5, 6),
                               min_child_weight = c(1, 3, 5))
phase1_results <- as.data.frame(t(apply(phase1_grid, 1, function(row) {
  res <- run_cv(modifyList(base_params,
                            list(max_depth        = row[["max_depth"]],
                                 min_child_weight = row[["min_child_weight"]])))
  c(row, best_auc = res$best_auc, best_nrounds = res$best_nrounds)
})))
print(phase1_results[order(-phase1_results$best_auc), ])

best1      <- phase1_results[which.max(phase1_results$best_auc), ]
best_depth <- best1$max_depth
best_mcw   <- best1$min_child_weight
cat(sprintf("\nBest: max_depth = %g | min_child_weight = %g | CV AUC = %.4f\n",
            best_depth, best_mcw, best1$best_auc))

# grid search visualization
p1_plot <- phase1_results
p1_plot$max_depth        <- factor(p1_plot$max_depth)
p1_plot$min_child_weight <- factor(p1_plot$min_child_weight)
print(
  ggplot(p1_plot, aes(x = max_depth, y = min_child_weight, fill = best_auc)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.3f", best_auc)), size = 3.5) +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "CV AUC") +
    geom_tile(data = subset(p1_plot,
                             max_depth        == as.character(best_depth) &
                             min_child_weight == as.character(best_mcw)),
              color = "red", fill = NA, linewidth = 1.2) +
    labs(title    = "XGBoost Phase 1 -- max_depth x min_child_weight",
         subtitle = sprintf("Best: max_depth = %g, min_child_weight = %g",
                            best_depth, best_mcw),
         x = "max_depth", y = "min_child_weight") +
    theme_minimal()
)

```

Best: max_depth = 3 | min_child_weight = 3 | CV AUC = 0.9010

Phase 2: subsample and colsample_bytree

```{r}
phase2_grid    <- expand.grid(subsample        = c(0.6, 0.8, 1.0),
                               colsample_bytree = c(0.6, 0.8, 1.0))
phase2_results <- as.data.frame(t(apply(phase2_grid, 1, function(row) {
  res <- run_cv(modifyList(base_params,
                            list(max_depth        = best_depth,
                                 min_child_weight = best_mcw,
                                 subsample        = row[["subsample"]],
                                 colsample_bytree = row[["colsample_bytree"]])))
  c(row, best_auc = res$best_auc, best_nrounds = res$best_nrounds)
})))
print(phase2_results[order(-phase2_results$best_auc), ])

best2    <- phase2_results[which.max(phase2_results$best_auc), ]
best_sub <- best2$subsample
best_cst <- best2$colsample_bytree
cat(sprintf("\nBest: subsample = %g | colsample_bytree = %g | CV AUC = %.4f\n",
            best_sub, best_cst, best2$best_auc))

# phase 2 visualization
p2_plot <- phase2_results
p2_plot$subsample        <- factor(p2_plot$subsample)
p2_plot$colsample_bytree <- factor(p2_plot$colsample_bytree)
print(
  ggplot(p2_plot, aes(x = subsample, y = colsample_bytree, fill = best_auc)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.3f", best_auc)), size = 3.5) +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "CV AUC") +
    geom_tile(data = subset(p2_plot,
                             subsample        == as.character(best_sub) &
                             colsample_bytree == as.character(best_cst)),
              color = "red", fill = NA, linewidth = 1.2) +
    labs(title    = "XGBoost Phase 2 -- subsample x colsample_bytree",
         subtitle = sprintf("Best: subsample = %g, colsample_bytree = %g",
                            best_sub, best_cst),
         x = "subsample", y = "colsample_bytree") +
    theme_minimal()
)
```

Best: subsample = 0.8 | colsample_bytree = 0.6 | CV AUC = 0.9055

Phase 3: gamma 

```{r}
phase3_results <- do.call(rbind, lapply(c(0, 0.1, 0.2, 0.3, 0.5), function(g) {
  res <- run_cv(modifyList(base_params,
                            list(max_depth        = best_depth,
                                 min_child_weight = best_mcw,
                                 subsample        = best_sub,
                                 colsample_bytree = best_cst,
                                 gamma            = g)))
  data.frame(gamma = g, best_auc = res$best_auc, best_nrounds = res$best_nrounds)
}))
print(phase3_results[order(-phase3_results$best_auc), ])

best_gamma <- phase3_results$gamma[which.max(phase3_results$best_auc)]
cat(sprintf("\nBest: gamma = %g | CV AUC = %.4f\n",
            best_gamma, max(phase3_results$best_auc)))

# phase 3 visualization
print(
  ggplot(phase3_results, aes(x = factor(gamma), y = best_auc, fill = best_auc)) +
    geom_col(color = "white") +
    geom_text(aes(label = sprintf("%.3f", best_auc)), vjust = -0.4, size = 3.5) +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "CV AUC") +
    geom_col(data = subset(phase3_results, gamma == best_gamma),
             color = "red", fill = NA, linewidth = 1.2) +
    labs(title    = "XGBoost Phase 3 -- gamma (split penalty)",
         subtitle = sprintf("Best: gamma = %g", best_gamma),
         x = "gamma", y = "CV AUC") +
    theme_minimal() + theme(legend.position = "none")
)
```

Best: gamma = 0.3 | CV AUC = 0.9093

Phase 4: eta and best rounds 

```{r}
phase4_results <- do.call(rbind, lapply(c(0.1, 0.05, 0.01), function(e) {
  max_rounds <- ifelse(e <= 0.01, 1000, 500)
  res <- run_cv(modifyList(base_params,
                            list(max_depth        = best_depth,
                                 min_child_weight = best_mcw,
                                 subsample        = best_sub,
                                 colsample_bytree = best_cst,
                                 gamma            = best_gamma,
                                 eta              = e)),
                nrounds = max_rounds)
  data.frame(eta = e, best_auc = res$best_auc, best_nrounds = res$best_nrounds)
}))
print(phase4_results[order(-phase4_results$best_auc), ])

best_eta     <- phase4_results$eta[which.max(phase4_results$best_auc)]
best_nrounds <- phase4_results$best_nrounds[which.max(phase4_results$best_auc)]
cat(sprintf("\nBest: eta = %g | nrounds = %d | CV AUC = %.4f\n",
            best_eta, best_nrounds, max(phase4_results$best_auc)))

# phase 4 visualization
print(
  ggplot(phase4_results, aes(x = factor(eta), y = best_auc, fill = best_auc)) +
    geom_col(color = "white") +
    geom_text(aes(label = sprintf("%.3f", best_auc)), vjust = -0.4, size = 3.5) +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "CV AUC") +
    geom_col(data = subset(phase4_results, eta == best_eta),
             color = "red", fill = NA, linewidth = 1.2) +
    labs(title    = "XGBoost Phase 4 -- eta (learning rate)",
         subtitle = sprintf("Best: eta = %g (nrounds = %d)", best_eta, best_nrounds),
         x = "eta", y = "CV AUC") +
    theme_minimal() + theme(legend.position = "none")
)

```

Best: eta = 0.1 | nrounds = 72 | CV AUC = 0.9093

Update final optimal hyperparameters

```{r}
final_params <- list(
  booster          = "gbtree",
  objective        = "binary:logistic",
  eval_metric      = c("error", "auc"),
  eta              = best_eta,
  max_depth        = best_depth,
  min_child_weight = best_mcw,
  subsample        = best_sub,
  colsample_bytree = best_cst,
  gamma            = best_gamma
)

print(as.data.frame(final_params[c("eta","max_depth","min_child_weight",
                                    "subsample","colsample_bytree","gamma")]))
```

Train the final model 

```{r}
cat(sprintf("\nTraining final model: nrounds = %d\n", best_nrounds))
xgb_model <- xgb.train(
  params        = final_params,
  data          = dtrain,
  nrounds       = best_nrounds,
  watchlist     = list(train = dtrain, eval = dtest),
  verbose       = 1,
  print_every_n = 10
)
```

Find the optimal classification threshold using Youden's J statistic

```{r}
# Get predicted probabilities for training data
pred_train_prob <- predict(xgb_model, newdata = dtrain)

# Find optimal threshold via Youden's J on training ROC
roc_train   <- roc(df_train$target, pred_train_prob, levels = c("no_disease", "disease"))
best_coords <- coords(roc_train, "best", best.method = "youden",
                      ret = c("threshold", "sensitivity", "specificity"))
opt_thresh  <- best_coords$threshold
cat(sprintf("Optimal threshold (Youden's J): %.4f\n", opt_thresh))
```

Optimal threshold (Youden's J): 0.4215. Using this threshold, we calculate train and test error.

```{r}
# Train error using optimal threshold
pred_train_label <- ifelse(pred_train_prob >= opt_thresh, "disease", "no_disease")
train_err <- mean(pred_train_label != as.character(df_train$target))
cat(sprintf("Train error: %.4f\n", train_err))

# Test error using the same threshold
xgb_probs <- predict(xgb_model, newdata = dtest)
xgb_preds <- ifelse(xgb_probs >= opt_thresh, "disease", "no_disease")
test_err  <- mean(xgb_preds != as.character(df_test$target))
cat(sprintf("Test error: %.4f\n", test_err))
```

Train error: 0.0588, test error: Test error: 0.1695.

Evaluation of model: 

```{r}
xgb_preds <- factor(xgb_preds, levels = c("no_disease", "disease"))
xgb_cm <- confusionMatrix(xgb_preds, df_test$target, positive = "disease")
print(xgb_cm)

xgb_roc <- roc(y_test, xgb_probs, quiet = TRUE)
cat(sprintf("Test AUC: %.4f\n", auc(xgb_roc)))

xgb_f1 <- xgb_cm$byClass["F1"]
cat(sprintf("F1 score: %.4f\n", xgb_f1))
```

Confusion Matrix visualization:
```{r}
xgb_cm_df <- as.data.frame(xgb_cm$table)
print(
  ggplot(xgb_cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), size = 7, fontface = "bold") +
    scale_fill_gradient(low = "lightyellow", high = "steelblue", name = "Count") +
    labs(title    = "XGBoost — Confusion Matrix",
         x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 13)
)
```

ROC Curve
```{r}
print(
  ggplot(data.frame(tpr = xgb_roc$sensitivities,
                    fpr = 1 - xgb_roc$specificities),
         aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    labs(title = "XGBoost — ROC Curve", x = "False Positive Rate", y = "True Positive Rate") +
    theme_minimal()
)
```

Variable Importance

```{r}
# renaming the variables for visualization purposes, = colnames(X_train)
feature_names <- c("Age", 
                   "Sex", 
                   "Chest pain: anginal",
                   "Chest pain: atypical angina", 
                   "Chest pain: typical angina",
                   "Resting blood pressure",
                   "Serum cholesterol",
                   "Fasting blood sugar >120 mg/dl",
                   "Resting electrocardiographic results: abnormal",
                   "Resting electrocardiographic results: hypertrophy",
                   "Maximum heart rate achieved",
                   "Exercise induced angina",
                   "ST depression",
                   "Slope: flat",
                   "Slope: down sloping",
                   "Number of major vessels", 
                   "Thal: fixed defect",
                   "Thal: reversible defect"
)

importance_mat <- xgb.importance(feature_names = feature_names, 
                                  model = xgb_model)
print(importance_mat)

xgb.plot.importance(importance_mat,
                    main = "XGBoost Feature Importance")
```

# Model Comparison

```{r}
comparison <- data.frame(
  Model       = c("Logistic Regression","XGBoost", "SVM (RBF)"),
  Test_AUC    = round(c(auc(lr_roc), 
                        auc(xgb_roc), 
                        auc(svm_roc)), 
                      4),
  Accuracy    = round(c(lr_cm$overall["Accuracy"],
                        xgb_cm$overall["Accuracy"],
                         svm_cm$overall["Accuracy"]), 
                      4),
  Accuracy_CI_Lower = round(c(lr_cm$overall["AccuracyLower"],
                        xgb_cm$overall["AccuracyLower"],
                         svm_cm$overall["AccuracyLower"]), 
                      4),
  Accuracy_CI_Upper = round(c(lr_cm$overall["AccuracyUpper"],
                        xgb_cm$overall["AccuracyUpper"],
                         svm_cm$overall["AccuracyUpper"]), 
                      4),
  Sensitivity = round(c(lr_cm$byClass["Sensitivity"],
                        xgb_cm$byClass["Sensitivity"],
                        svm_cm$byClass["Sensitivity"]), 
                      4),
  Specificity = round(c(lr_cm$byClass["Specificity"], 
                        xgb_cm$byClass["Specificity"],
                        svm_cm$byClass["Specificity"]), 
                      4),
  F1          = round(c(lr_f1, xgb_f1, svm_f1), 4)
)
print(comparison, row.names = FALSE)

# ROC curve overlay
plot(lr_roc, col = "steelblue", lwd = 2, legacy.axes = TRUE,
     main = "ROC Curves — Logistic, XGBoost, SVM (Test Set)")
plot(xgb_roc, col = "forestgreen", lwd = 2, add = TRUE, legacy.axes = TRUE)
plot(svm_roc, col = "tomato", lwd = 2, add = TRUE, legacy.axes = TRUE)
legend("bottomright",
       legend = c(
         sprintf("Logistic  (AUC = %.3f)", auc(lr_roc)),
         sprintf("XGBoost  (AUC = %.3f)", auc(xgb_roc)),
         sprintf("SVM RBF  (AUC = %.3f)", auc(svm_roc))),
       col = c("steelblue", "forestgreen", "tomato"), lwd = 2, bty = "n")
```
